---
title: "Karatina University"
author: "John Karuitha"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
params:
  course: 'BBM 450: Practical Accounting Data Processing'
subtitle: School of Business
always_allow_html: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## `r str_to_upper(params$course)`

## COURSE PURPOSE
To introduce students to methods for finding, transferring and processing existing data sources. The course emphasizes practical approaches to pre-statistical data processing and analysis with statistical software on a PC with a MS Windows operating system, work very similarly. 

## COURSE LEARNING OUTCOMES

i. Convert a raw data set to an analytic data set.
ii. Manipulate variety of existing data resources potentially useful in accounting systems
iii. Use existing data conversion in software. 
iv. Utilize complex sampling and data processing designs. 

## COURSE CONTENT

1. Introduction to accounting information processing concepts. 

2. Accounting information systems development and management. 

3. Application of information systems in accounting reports. 

4. Accounting software and end-user software. 

5. Practical with latest Accounting packages. 

   - Ms Excel- Beginner to Advanced.
   - Quick books (supplemented by GNU Cash).
   
## MODE OF DELIVERY OF THE COURSE
The course will be delivered using physical and on-line lectures and discussions based on case studies and presentations by students.

## INSTRUCTIONAL MATERIALS/ EQUIPMENT

- Audio-visual devices.
- Ms Excel Software.
- GNU Cash Software.
- computers/internet services.
- journals, newspapers.
- chalk/pens and white boards.
- flip charts.

## About GNU Cash

GNU cash is an open source accounting software. Please search on the internet if you need to install the software. It is available for free.

## COURSE ASSESSMENT
The course will be assessed using 3 practical continuous assessment tests, 1 project-based class assignment and 1 final exam, as shown in Table 1 below. Note that all assessments will be practical. I will especially put emphasis on Ms Excel as it is a core tool for accountants 

useful texts for the course include @proctor2004building, @mayes2020financial, @hall2015accounting, as well as the numerous resources available for free online. 

```{r, echo = FALSE}
library(tibble)
library(kableExtra)
tribble(~ Assessment, ~ Contribution,
        "CAT 1: Practical", "5%",
        "CAT 2: Practical", "5%",
        "CAT 3: Sit in: Practical", "10%",
        "CAT 4: Group Presentations", "10%", 
        "Final Practical Examination", "70%", 
        "TOTAL", "100%") %>% 
  
  kableExtra::kbl(., caption = "Course Evaluation", 
      
  booktabs = TRUE) %>% 
  
  kable_paper(full_width = FALSE, font_size = 12,
              
              latex_option = "hold_position")
```

## Class Times

We shall hold the class between 8.00 am and 11.00 am in the business school computer lab. 

## Contacts

You can reach me on `jkaruitha@karu.ac.ke`.

## References






