---
subtitle: "**COURSE OUTLINE**"
author: "*John Karuitha*, jkaruitha@karu.ac.ke"
date: "`r format(Sys.Date(), format = '%A %B %d, %Y')`"
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{center}
    \includegraphics[width=2in,height=2in]{karu2.png}\LARGE\\}
  - \posttitle{\end{center}}
title: "**BBM 419 : Financial Statements Analysis**"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(kableExtra)
```

## **Course Purpose**

The course will assist the learner to analyze financial statements as a basis for extracting information for business decision making, especially in investments.

## **Learning Outcomes**

By the end of the course, the learner should:

i. Identify the various sources of information for the analysis of a business.
ii. Analyse a business using the financial statement analysis tools and make decisions.

## **Course Content**

1. Introduction to financial statements analysis.
2. The financial report and financial statements.

- Components of a financial report.
- GAAPs: Assumptions and principles of accounting.
- Accounting standards (IFRS). 

3. Using Financial Statement Information.

- Why evaluate financial statements? 
- Internal uses, external uses.
- Choosing a benchmark.
- Merits and limitations of financial statements analysis.

2. Getting comfortable with Ms Excel for Financial Analysis (Week 5 to 7).
4. Cash flows and taxes.
5. Standardized Financial Statements.

- Vertical analysis
- Horizontal analysis/ Base year analysis
- Combined vertical and horizontal analysis.

6. Ratio Analysis & the Du Point Identity.
7. Earnings analysis.
8. Cash flow analysis.
9. Time trend analysis & peer group analysis.

## **Teaching Methodology**

- Lectures and discussions.
- Case studies.
- Class discussions and presentations.
- Instructional Materials and Equipment.
- Computers/internet services, journals, newspapers, chalk/pens and white boards.

## **Instructional Materials and Equipment**

- Prescribed textbooks.
- Computers/internet services.
- Journal and newspaper articles.

## **Course Assessment**

```{r}
tribble(~ Asessment, ~ Score,
        
        "Continuous Assessments", "10%",
        
        "Ms Excel Practical Assessment", "10%",
        
        "Classroom assignments",  "10%",
        
        "End of semester examinations",  "70%",
        
        "TOTAL", "100%"
        
        ) %>% kbl(., booktabs = TRUE) %>% 
        
        kable_classic(full_width = TRUE, latex_options = "striped")
```

## **References**

Subramanyam, K. (2014). *Financial Statements Analysis, 11th Edition*. London. McGraw          Hill Education. 

Proctor, K.S. (2009). *Building Financial Models with Microsoft Excel: A Guide for            Business Professionals*. New York. Wiley.

Fridson, M., & Alvarez, F. (2002). *Financial Statement Analysis: A Practitioners             Guide*. Hoboken, New Jersey: John Wiley & Sons, Inc.

Robinson, T., & Henry, E. (2009). *International Financial Statements Analysis*.               Hoboken, New Jersey: John Wiley & Sons, Inc.