---
title: "Karatina University"
author: "John Karuitha"
date: "`r format(Sys.Date(), '%A, %B %d, %Y')`"
output:
  pdf_document: default
  html_document: default
  word_document: default
params:
  course: 'BHS 220: HEALTH CARE FINANCIAL MANAGEMENT'
subtitle: School of Business
always_allow_html: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `r params$course`

## COURSE PURPOSE
This course will explore the use of financial information essential to the operational management of Health Services, including estimating costs and profits, planning and budgeting, using metrics to monitor operations, and working with financial statements. The main reference text books are @ward2015health, @tirole2010theory, and Kenya Public Financial Management Act, 2012 [@pfma2012]. 

## COURSE LEARNING OUTCOMES

Upon successful completion of this course, students should be able to:

1. Explain the procedure for managing working capital in a health facility.
2. Explain the importance of financial auditing in the effective management of health organizations.
3. Apply basic concepts of forecasting and budgeting effectively in a health facility.
4. Describe the Kenyan legal and regulatory framework for managing public finances.

## COURSE CONTENT

1. Introduction to Financial Management.

   - What is Finance?
   - Definition of Financial Management.
   - What Do Finance Managers do?
   - Objectives of Financial Management. 
   - Agency Theory.
        
2. Concepts of Health Care Financial Management.

   - Features of Healthcare financing.
   
3. Accounting.

   - Definitions.
   - Distinguishing Accounting from Finance.
   - Branches of Accounting.
   - Financial Accounting Framework.
   - Components of a Financial Report.
   - Depreciation. 

4. Auditing.

   - Definition.
   - Importance of Auditing.
   - Types of Audits.
   - Audit Evidence.
   
3. Cost-Volume-Profit (CVP) Analysis.

   - Cost and Cost Behaviour.
   
        - Fixed Costs.
        - Variable costs.
        - Mixed Costs.
        
   - Applications: Break Even Analysis.
  
5. Analysis of Health Care Financial Statements.

   - Overview of Financial Statements Analysis.
   - Common Size Financial Statements.
   - Ratio Analysis. 

7. Budgeting Responsibility accounting.

   - Budgeting Process.
   - Cash Budget.

8. Capital Budgeting. 

   - Capital Budgeting Process.
   - Introduction to capital budgeting techniques.

6. Working Capital Management.

10. Constitutional and legal aspects of public finance control in Kenya

   - Introduction.
   - The Public Financial Management Act.

## MODE OF DELIVERY OF THE COURSE
The course will be delivered using physical and on-line lectures and discussions based on case studies and presentations by students.

## INSTRUCTIONAL MATERIALS/ EQUIPMENT

- Audio-visual devices.
- Computers/internet services.
- Journals, newspapers.
- Flip charts.

## COURSE ASSESSMENT
The course will be assessed using 3 continuous assessment tests, 1 project-based class assignment and 1 final exam, as shown in Table 1 below.

```{r, echo = FALSE}
library(tibble)
library(kableExtra)
tribble(~ Assessment, ~ Contribution,
        "CAT 1: Multiple Choice", "5%",
        "CAT 2: Multiple Choice", "5%",
        "CAT 3: Sit in: Essays and Computations", "10%",
        "CAT 4:Group Presentations", "10%", 
        "Final Examination", "70%", 
        "TOTAL", "100%") %>% 
  
  kableExtra::kbl(., caption = "Course Evaluation", 
      
  booktabs = TRUE) %>% 
  
  kable_paper(full_width = FALSE, font_size = 12,
              
              latex_option = "hold_position")
```


## References






